---
title: "Marthas Vineyard SET report"
subtitle: "Produced for Martha’s Vineyard Commission" 
date: "`r Sys.Date()`"
author:
  - name: Laura Feher
    affiliations: 
    - name: NPS NCBN I&M Network
    email: laura_feher@nps.gov
    orcid: 0000-0002-5983-6190
bibliography: references.bib
format: 
  html:
    code-links:
      - text: "Github repo"
        icon: file-code
        href: "https://github.com/laura-feher/SETs_Marthas_Vineyard"
css: style.css
toc: true 
code-fold: true
code-overflow: wrap
crossref: 
  title-delim: "."
embed-resources: true
execute:
  message: false
  warning: false
  echo: false
editor: visual
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(here)
library(ggthemr)
library(ggedit)
library(ggtext)

knitr::opts_chunk$set(
    out.width = "70%", # enough room to breath
    fig.width = 4,     # reasonable size
    fig.asp = 0.618   # golden ratio
)

map(list.files(here::here("R"), pattern = "\\.R$", full.names = TRUE), function(x)
  source(file = x, local = knitr::knit_global()))

aquinnah <- readRDS(here::here("data", "aquinnah.rds"))

felix_neck <- readRDS("C:/Users/lfeher/OneDrive - DOI/SETs_Marthas_Vineyard/data/felix_neck.rds")

woods_hole_slr <- readRDS("C:/Users/lfeher/OneDrive - DOI/SETs_Marthas_Vineyard/data/woods_hole_slr.rds")
```

**NOTE:** The SET-MH approach has been widely used for documenting and interpreting trends in surface elevation dynamics over time and assessing wetland vulnerability to sea-level rise. Thus, the methods used here for the analyses of SET and sea-level rise data follow the standard conventions outlined in the following peer reviewed literature: @cahoon2002a; @cahoon2002b; @cahoon2006; @cahoon2024; @callaway2013; @lynch2015.

## Methods: Data Analyses

Prior to data analyses, we subtracted the elevation reading of each pin on each sampling date from its initial value to determine cumulative elevation change [@lynch2015]. The pin-level elevation readings for each measurement date were then averaged across each of the four bearings on each benchmark, then the bearing-level values were averaged to the plot level, and finally the plot-level values were averaged to the site level. To estimate the rate of marsh surface elevation change at each site, we used a simple linear regression model where the independent variable was the time since the first measurement and the dependent variable was site-level cumulative marsh surface elevation change as outlined in @feher2024 and @russell2022.

## Methods: Relative Sea-Level Rise

For the long-term rate of relative sea-level rise (RSLR; also known as local sea-level rise), we obtained the rate published by NOAA for the nearby tide gauge at `r woods_hole_slr$woods_hole_slr_long$Station.Name` [@NOAAtidesWH] for the entire period of record from `r woods_hole_slr$woods_hole_slr_long$First.Year` to `r woods_hole_slr$woods_hole_slr_long$Last.Year`. We estimated a recent rate of RSLR using a linear regression applied to the monthly mean sea-level data with the average seasonal cycle removed by subsetting the data from the Woods Hole gauge for the most recent tidal epoch (`r year(unique(woods_hole_slr$woods_hole_slr_recent$data[[1]]$min_date))` -`r year(unique(woods_hole_slr$woods_hole_slr_recent$data[[1]]$max_date))`).

We estimated future projected RSLR rates using the five alternative SLR scenarios produced by @sweet2022 for the 5th National Climate Assessment. The Low, Intermediate-Low, Intermediate, Intermediate-High, and High SLR scenarios correspond respectively to global mean sea level increases of 0.3 m, 0.5 m, 1.0 m, 1.5 m, and 2.0 m by 2100. To approximate the RSLR rate by 2100 under each of the five scenarios, we used the data provided by @sweet2022 to estimate future RSLR rates for the 2090-2100 decade from the Woods Hole RSLR projections.

## Results: Rates of Marsh Surface Elevation Change

The rate of marsh surface elevation change at Aquinnah was `r aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate)` mm/yr (*r*^2^ = `r aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% pull(R2)`, *p* `r aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% mutate(p = format_pval(p.value, output_type = "subtitle")) %>% pull(p)` ) for the period between `r year(first(aquinnah$aq_cumu_site$date))` to `r year(last(aquinnah$aq_cumu_site$date))` ([Fig. @fig-plot-aq-simple]). The rate of marsh surface elevation change at Felix Neck was `r felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate)`mm/yr (*r*^2^ = `r felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% pull(R2)`, *p* `r felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% mutate(p = format_pval(p.value, output_type = "subtitle")) %>% pull(p)` ) for the period between `r year(first(felix_neck$fn_cumu_site$date))` to `r year(last(felix_neck$fn_cumu_site$date))` ([Fig @fig-plot-fn-simple]).

```{r}
#| label: fig-plot-aq-simple
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("Marsh surface elevation change at ", unique(aquinnah$aq_cumu_site$site), " between ", year(first(aquinnah$aq_cumu_site$date)), "-", year(last(aquinnah$aq_cumu_site$date)), ".")

mv_plot(site_df = aquinnah$aq_cumu_site, site_rate_df = aquinnah$aq_linear_mod_site, plot_type = "simple SET plot")
```

```{r}
#| label: fig-plot-fn-simple
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("Marsh surface elevation change at ", unique(felix_neck$fn_cumu_site$site), " between ", year(first(felix_neck$fn_cumu_site$date)), "-", year(last(felix_neck$fn_cumu_site$date)), ".")

mv_plot(site_df = felix_neck$fn_cumu_site, site_rate_df = felix_neck$fn_linear_mod_site, plot_type = "simple SET plot")
```

## Results: Marsh Surface Elevation Change vs. Long-term & Recent SLR

The long-term rate of RSLR at the nearby Woods Hole tide gauge was `r woods_hole_slr$woods_hole_slr_long$MSL.Trends.mm.yr.` ± `r woods_hole_slr$woods_hole_slr_long$Perc.95.CI.mm.yr.` mm/yr, whereas the recent rate of RSLR was `r woods_hole_slr$woods_hole_slr_recent %>% filter(term == "yr") %>% mutate(recent_slr = paste0(format_result_vals(estimate), " ± ", format_result_vals(std.error))) %>% pull(recent_slr)` mm/yr. Accordingly, while the marshes at both Aquinnah and Felix Neck have been building surface elevation at a rate close to or slightly above the long-term rate of SLR, neither marsh is keeping pace with the accelerated recent rate of sea-level rise in the region ([Fig @fig-plot-aq-slr1]/[Fig @fig-plot-aq-slr2] and [Fig @fig-plot-fn-slr1]/[Fig @fig-plot-fn-slr2]).

```{r}
#| label: fig-plot-aq-slr1
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. A) Marsh surface elevation change at Aquinnah as compared to long-term and recent rates of SLR. Values to the right of the colored lines represent the rate of marsh surface elevation change at Aquinnah and the long-term and recent rates of SLR from the Woods Hole gauge.")

mv_plot(site_df = aquinnah$aq_cumu_site, site_rate_df = aquinnah$aq_linear_mod_site, plot_type = "SLR labeled")
```

```{r}
#| label: fig-plot-aq-slr2
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. B) Marsh surface elevation change at Aquinnah as compared to long-term and recent rates of SLR. The rate of surface elevation change at Aquinnah between ", year(first(aquinnah$aq_cumu_site$date)), "-", year(last(aquinnah$aq_cumu_site$date)), " was ", aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate), " mm/yr (", "<i>r^2^</i>", " = ", aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% pull(R2), ", ", "<i>p</i>", aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% mutate(p = format_pval(p.value, output_type = "subtitle")) %>% pull(p), "). Using the data published by NOAA for the nearby gauge at Woods Hole, MA, the long-term rate of SLR was ", paste0(woods_hole_slr$woods_hole_slr_long$MSL.Trends.mm.yr., " ± ", woods_hole_slr$woods_hole_slr_long$Perc.95.CI.mm.yr.), " mm/yr and the recent rate of SLR was ", woods_hole_slr$woods_hole_slr_recent %>% filter(term == "yr") %>% mutate(rate = paste0(format_result_vals(estimate), " ± ", format_result_vals(std.error))) %>% pull(rate), " mm/yr.")

mv_plot(site_df = aquinnah$aq_cumu_site, site_rate_df = aquinnah$aq_linear_mod_site, plot_type = "SLR no labels")
```

```{r}
#| label: fig-plot-fn-slr1
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. A) Marsh surface elevation change at Felix Neck as compared to long-term and recent rates of SLR. Values to the right of the colored lines represent the rate of marsh surface elevation change at Felix Neck and the long-term and recent rates of SLR from the Woods Hole gauge.")

mv_plot(site_df = felix_neck$fn_cumu_site, site_rate_df = felix_neck$fn_linear_mod_site, plot_type = "SLR labeled")
```

```{r}
#| label: fig-plot-fn-slr2
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. B) Marsh surface elevation change at Felix Neck as compared to long-term and recent rates of SLR. The rate of surface elevation change at Felix Neck between ", year(first(felix_neck$fn_cumu_site$date)), "-", year(last(felix_neck$fn_cumu_site$date)), " was ", felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate), " mm/yr (", "<i>r^2^</i>", " = ", felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% pull(R2), ", ", "<i>p</i>", felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% mutate(p = format_pval(p.value, output_type = "subtitle")) %>% pull(p), "). Using the data published by NOAA for the nearby gauge at Woods Hole, MA, the long-term rate of SLR was ", paste0(woods_hole_slr$woods_hole_slr_long$MSL.Trends.mm.yr., " ± ", woods_hole_slr$woods_hole_slr_long$Perc.95.CI.mm.yr.), " mm/yr and the recent rate of SLR was ", woods_hole_slr$woods_hole_slr_recent %>% filter(term == "yr") %>% mutate(rate = paste0(format_result_vals(estimate), " ± ", format_result_vals(std.error))) %>% pull(rate), " mm/yr.")

mv_plot(site_df = felix_neck$fn_cumu_site, site_rate_df = felix_neck$fn_linear_mod_site, plot_type = "SLR no labels")
```

## Results: Marsh Surface Elevation Change vs. Future SLR

Future predicted rates of RSLR for the Woods Hole gauge were `r woods_hole_slr$woods_hole_slr_future %>% pull(rsl_rate_mm_yr) %>% knitr::combine_words(.)` mm/yr for the low, intermediate-low, intermediate, intermediate-high, and high sea-level rise scenarios, respectively, for the decade between 2090 to 2100. Although the marsh at Aquinnah has been building surface elevation at a rate close to or slightly above the rate of SLR predicted by the future low scenario (`r aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate)` mm/yr), the rate of marsh surface elevation change at the site was lower than any of the other future predicted SLR scenarios. Thus, its unlikely that the marsh at Aquinnah will be able to keep pace with future predicted rates of sea-level rise that are any greater than the lowest SLR scenario ([Fig @fig-plot-aq-future-slr1]/[Fig @fig-plot-aq-future-slr2]). Similarly, the rate of marsh surface elevation change at Felix Neck was lower than all of the future predicted SLR scenarios (`r felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate)` mm/yr), making it unlikely that this marsh will be able to keep pace with any future acceleration in the rate of SLR ([Fig @fig-plot-fn-future-slr1]/[Fig @fig-plot-fn-future-slr2]).

```{r}
#| label: fig-plot-aq-future-slr1
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. A) Marsh surface elevation change at Aquinnah as compared to future predicted rates of SLR. Values to the right of the colored lines represent the rate of marsh surface elevation change at Aquinnah and the rates of sea-level rise predicted by each of the future SLR scenarios for the nearby gauge at Woods Hole, MA.")

mv_plot(site_df = aquinnah$aq_cumu_site, site_rate_df = aquinnah$aq_linear_mod_site, plot_type = "future SLR labeled")
```

```{r}
#| label: fig-plot-aq-future-slr2
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. B) Marsh surface elevation change at Aquinnah as compared to future predicted rates of SLR. The rate of surface elevation change at Aquinnah between ", year(first(aquinnah$aq_cumu_site$date)), "-", year(last(aquinnah$aq_cumu_site$date)), " was ", aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate), " mm/yr (", "<i>r^2^</i>", " = ", aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% pull(R2), ", ", "<i>p</i>", aquinnah$aq_linear_mod_site %>% filter(term == "date_num") %>% mutate(p = format_pval(p.value, output_type = "subtitle")) %>% pull(p), "). Using the data published by NOAA for the nearby gauge at Woods Hole, MA, future predicted rates of SLR were ", woods_hole_slr$woods_hole_slr_future %>% pull(rsl_rate_mm_yr) %>% knitr::combine_words(.), " mm/yr for the low, intermediate-low, intermediate, intermediate-high, and high SLR scenarios, respectively.")

mv_plot(site_df = aquinnah$aq_cumu_site, site_rate_df = aquinnah$aq_linear_mod_site, plot_type = "future SLR no labels")
```

```{r}
#| label: fig-plot-fn-future-slr1
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. A) Marsh surface elevation change at Felix Neck as compared to future predicted rates of SLR. Values to the right of the colored lines represent the rate of marsh surface elevation change at Felix Neck and the rates of sea-level rise predicted by each of the future SLR scenarios for the nearby gauge at Woods Hole, MA.")

mv_plot(site_df = felix_neck$fn_cumu_site, site_rate_df = felix_neck$fn_linear_mod_site, plot_type = "future SLR labeled")
```

```{r}
#| label: fig-plot-fn-future-slr2
#| fig-cap: !expr lab
#| eval.after: 'fig.cap'

lab <- paste0("(Opt. B) Marsh surface elevation change at Felix Neck as compared to future predicted rates of SLR. The rate of surface elevation change at Felix Neck between ", year(first(felix_neck$fn_cumu_site$date)), "-", year(last(felix_neck$fn_cumu_site$date)), " was ", felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% pull(linear_rate), " mm/yr (", "<i>r^2^</i>", " = ", felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% pull(R2), ", ", "<i>p</i>", felix_neck$fn_linear_mod_site %>% filter(term == "date_num") %>% mutate(p = format_pval(p.value, output_type = "subtitle")) %>% pull(p), "). Using the data published by NOAA for the nearby gauge at Woods Hole, MA, future predicted rates of SLR were ", woods_hole_slr$woods_hole_slr_future %>% pull(rsl_rate_mm_yr) %>% knitr::combine_words(.), " mm/yr for the low, intermediate-low, intermediate, intermediate-high, and high SLR scenarios, respectively.")

mv_plot(site_df = felix_neck$fn_cumu_site, site_rate_df = felix_neck$fn_linear_mod_site, plot_type = "future SLR no labels")
```
